---
title: "Calculate asymmetry of faces"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: facefuns.bib
csl: apa-numeric-superscript.csl
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Calculate asymmetry of faces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**PROCEED WITH CAUTION... WORK IN PROGRESS**

```{r setup}
library(facefuns)
library(ggplot2)
```

## Load and prep data
```{r}
path_to_tem <- system.file("extdata", "tem", package="facefuns")
remove_points <- c(45:50, 100:104, 116:125, 146:158, 159:164, 165:170, 171:174, 175:179, 184:185)

data <- read_lmdata(lmdata = path_to_tem,
                    remove_points = remove_points,
                    plot = FALSE)

shapedata <- facefuns2d(data = data,
                        pc_criterion = "broken_stick")
```

## Calculate asymmetry

*Needs tidying up. For now we have:*

* `calc_as`: calculates asymmetry as the Procrustes distance between each original template and its symmetrized counterpart[@komori2009]; does not correct for directional asymmetry (DA)

* `calc_fa`: calculates fluctuating asymmetry (FA) by correcting asymmetry as calculated above for DA following Klingenberg (2015)[@klingenberg2015]; DA is calculated as the average of original templates minus the average of mirrored templates

* `calc_fageo`: calculates FA, here using `geomorph::bilat.symmetry`. I am not entirely sure what  `bilat.symmetry` is doing under the hood and hence whether I'm implementing this correctly. Resulting scores are strongly correlated with the other two, but of course that doesn't mean much

```{r}
as <- calc_as(shapedata, mirroredlandmarks)
fa <- calc_fa(shapedata, mirroredlandmarks)
fa.geo <- calc_fageo(shapedata, mirroredlandmarks)

compare <- fa %>%
  dplyr::left_join(fa.geo, by="id") %>%
  dplyr::left_join(as, by="id") %>%
  # one picture is very asymmetric (face turned sideways) and was
  # excluded, so as to not inflate correlations between different scores
  dplyr::filter(id != "139") %>%
  dplyr::mutate(fa.geo = fa.geo/2)

as <- compare %>%
  tidyr::pivot_longer(!c(id, asym),
                      names_to = "other_fa_types",
                      values_to = "score") %>%
  ggplot(aes(x = asym, y = score, colour = other_fa_types)) +
  geom_point()

fa <- compare %>%
  tidyr::pivot_longer(!c(id, fa),
                      names_to = "other_fa_types",
                      values_to = "score") %>%
  ggplot(aes(x = fa, y = score, colour = other_fa_types)) +
  geom_point()

cowplot::plot_grid(as, fa, ncol = 2)
```



## References
