---
title: "Calculate sexual dimorphism of faces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculate sexual dimorphism of faces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**WORK IN PROGRESS**

[Eventually, this function should take as arguments: set of faces for which SD is to be calculated, set of male faces, set of female faces, type of calculation (vector/discriminant)]

This tutorial will explain how to calculate facial sexual dimorphism (SD) in a set of 2D faces, using (1) a vector method (REF) and (2) linear discriminant analyses (REF).

[Explain what these are/how they differ]

Vector scores can be calculated within- or out-of-set. For example, the sample for which you want to calculate sexual dimorphism may be small, but you might have a more sizable sample of female/male faces at hand from another project. You could use that bigger sample to construct your female-male vector, and then use it to calculate SD of the faces in the smaller set.

For this tutorial, we'll stay within set.

```{r setup}
library(facefuns)
```

## Read and prepare landmark data

[For now, start from PC scores; eventually make it so you can go directly from raw landmark data to scores?]

```{r}
path_to_tem <- system.file("extdata", "tem", package="facefuns")
remove_points <- c(45:50, 100:104, 116:125, 146:158, 159:164, 165:170, 171:174, 175:179, 184:185)

data <- read_shapedata(shapedata = path_to_tem,
                       remove_points = remove_points,
                       plot = FALSE)

shapedata <- quickstart(data = data,
                         rotate = "rotateC",
                         plot_sample = TRUE,
                         pc_criterion = "broken_stick")
```

We also need to know which of the faces are female/male.

```{r eval=FALSE}
data(LondonSet_info)

fem_ids <- LondonSet_info %>%
  dplyr::filter(face_sex == "female") %>%
  dplyr::pull(face_id)

mal_ids <- LondonSet_info %>%
  dplyr::filter(face_sex == "male") %>%
  dplyr::pull(face_id)

fem <-shapedata$pc_scores %>%
  as.data.frame() %>%
  dplyr::filter(grepl(paste(fem_ids, collapse="|"), shapedata$pc_scores$id)) %>%
  tibble::column_to_rownames("id")

mal <-shapedata$pc_scores %>%
  as.data.frame() %>%
  dplyr::filter(grepl(paste(mal_ids, collapse="|"), shapedata$pc_scores$id)) %>%
  tibble::column_to_rownames("id")


```

```{r eval=FALSE}
fem.mean <- t(as.matrix(colMeans(fem)))
mal.mean <- t(as.matrix(colMeans(mal)))

temp <- mal.mean - fem.mean
norm.dist <- sqrt(sum(temp ^ 2))
norm.vec <- temp / norm.dist

R <- nrow(PC.scores)
pc.matrix <- PC.scores %>% dplyr::select(starts_with("PC")) %>% as.matrix()
sd.vector <- numeric(R)
for (i in 1:R) {
  z1 <- pc.matrix[i, ] - fem.mean
  z2 <- z1 / norm.dist
  z3 <- z2 * norm.vec
  sd.vector[i] <- sum(z3)
}
```


